---
title: "RNA2Biomass_Constraint"
author: "Rachel L. Harris"
date: "2024-01-18"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r packages}
#Set working directory 
# setwd("~/Path_to/Harris2024_MarsLowP") # update to source directory

pkgTest <- function(...) {
  # Install dependency libraries if not installed and load them
  dep.list <- c(...)
  new_pkgs <- dep.list[!(dep.list %in% installed.packages()[,"Package"])]
  if(length(new_pkgs)) install.packages(new_pkgs)
  invisible(lapply(dep.list, library, character.only=T))
}
# Load dependencies and setup environment
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(reshape2)
library(dplyr)
library(plyr)
library(readxl)
library(ggthemes)
library(grid)
library(gtable)
library(gridExtra)
library(ggpubr)
library(scales)
library(ggthemes)
library(tidyverse)
library(tidyr)
library(drc)
library(stringr)
library(ggforce)
```

```{r, summarySE function}

## Gives count, mean, standard deviation, standard error of the mean, and confidence interval (default 95%).
##   data: a data frame.
##   measurevar: the name of a column that contains the variable to be summariezed
##   groupvars: a vector containing names of columns that contain grouping variables
##   na.rm: a boolean that indicates whether to ignore NA's
##   conf.interval: the percent range of the confidence interval (default is 95%)
summarySE <- function(data=NULL, measurevar, groupvars=NULL, na.rm=FALSE,
                      conf.interval=.95, .drop=TRUE) {
    library(plyr)

    # New version of length which can handle NA's: if na.rm==T, don't count them
    length2 <- function (x, na.rm=FALSE) {
        if (na.rm) sum(!is.na(x))
        else       length(x)
    }

    # This does the summary. For each group's data frame, return a vector with
    # N, mean, and sd
    datac <- ddply(data, groupvars, .drop=.drop,
      .fun = function(xx, col) {
        c(N    = length2(xx[[col]], na.rm=na.rm),
          mean = mean   (xx[[col]], na.rm=na.rm),
          sd   = sd     (xx[[col]], na.rm=na.rm)
        )
      },
      measurevar
    )

    # Rename the "mean" column    
    datac <- rename(datac, c("mean" = measurevar))

    datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean

    # Confidence interval multiplier for standard error
    # Calculate t-statistic for confidence interval: 
    # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
    ciMult <- qt(conf.interval/2 + .5, datac$N-1)
    datac$ci <- datac$se * ciMult

    return(datac)
}
```


```{r, read in data}
#Harris et al., 2021
RNA.2.Biomass_TableS9 <- as.data.frame(read_excel("Table S9.xlsx", sheet = 1, range = "A4:M34"), colnames = TRUE)

#This study
RNA.2.Biomass_TableS10 <- (as.data.frame(read_excel("Table S10_knowns.xlsx", sheet = 1, range = "A4:M34"), colnames = TRUE))
```


1) We have direct cell counts that we can map to RNA recovery for Harris et al., 2021 (https://doi.org/10.1038/s41598-021-91882-0)
2) We will fit a translated, log-transformed Michaelis-Menten regression to the relationship between recovered RNA and total cell counts, considering cell counts as the "responsible variable" and recovered RNA as the "explanatory variable"
```{r, regression modeling}

#to avoid negative log values for a regression fit, we will first translate the data by adding a +1 to all RNA concentration values #before taking the log10 transformation
logPlus1.RNA_long <- log10(RNA.2.Biomass_TableS9$`RNA (ng/µL)` + 1)
log.Cells_long <- log10(RNA.2.Biomass_TableS9$`Total Direct Cell Counts (AO) Experiment End`)

#build a data frame from the transformed data
log.data <- data.frame(logPlus1.RNA_long, log.Cells_long)

#Michaelis-Menten model fit of transformed data 
log.log.model_long <- drm(log.Cells_long~logPlus1.RNA_long, data = log.data, fct = MM.2())

summary(log.log.model_long) #assess goodness of model fit
conf <- confint(log.log.model_long, level = 0.95) #95% confidence interval

#save model parameters
Cmax <- log.log.model_long$fit$par[1] #functionally equivalent to Vmax, this is the maximum supportable cell concentration
Km<- log.log.model_long$fit$par[2] #Michaelis constant, representing the RNA concentration at which cell concentration is half
                                  # of its maximum

#get a list of predicted cell counts
Harris2021_PredCellCounts <- 10^fitted(log.log.model_long)

#get a list of residuals
res <- resid(log.log.model_long)

#Append predictions to RNA.2.Biomass_TableS9 data frame (this will be useful later)
RNA.2.Biomass_TableS9$PredCellCounts <- Harris2021_PredCellCounts

#compare probability distributions by building a quantile-quantile data frame
QQ <- data.frame(fitted(log.log.model_long), res)
colnames(QQ) <- c("Fitted Model Value", "Residual")

#model statistics
QQ$Residual_squared <- QQ$Residual^2
RSS <- sum(QQ$Residual_squared) #residual sum of squares
QQ$Observed <- log.Cells_long
TSS <- sum((QQ$Observed - mean(QQ$Observed)))^2 #total sum of squares

#produce a residual vs. fitted plot
plot.res.vs.fit <- ggplot(QQ, aes(x = `Fitted Model Value`, y = Residual))+
  geom_point()+
  geom_hline(yintercept = 0)+
  theme_base()

plot.res.vs.fit

#create a Q-Q plot for residuals
# A michaelis-menten fit is a reasonable regression
plot.qq <- ggplot(QQ, aes(sample = res))+
  stat_qq()+
  stat_qq_line()+
  theme_base()

plot.qq

```

```{r, model predictions}

#write a function that will ignore NAs in the dataset
my.max <- function(x) ifelse( !all(is.na(x)), max(x, na.rm=T), NA)

#now use the Michaelis-Menten relationship to constrain first order cell counts from recovered RNA
#treat RNA.ng.µL as variable S in the Michaelis-Menten equation and predict the sequence out 5 units past my.max
logmodel.mml <- data.frame(log10_RNA.ng.µL = seq(0.01, my.max(logPlus1.RNA_long)+5, 0.01))

#Treat Cell.count.constraint as variable v in the Michaelis-Menten equation
logmodel.mml$Cell.count.constraint <- predict(log.log.model_long, newdata = logmodel.mml, interval = "confidence", level = 0.95)

# Split "Cell.count.constraint" into "Predicted Total Cell Count", "Lower Bound", and "Upper Bound"
logmodel.mml[c('Predicted Total Cell Count', 'Lower Bound', 'Upper Bound')] <- str_split_fixed(logmodel.mml$Cell.count.constraint, ' ', 3)

# Rearrange columns and remove original Cell.count.constraint column
logmodel.mml <- logmodel.mml[c(1, 3:5)]
#turn characters back into numeric values
logmodel.mml$log10_RNA.ng.µL <- as.numeric(logmodel.mml$log10_RNA.ng.µL)
#predict cell counts by taking the exponent to the power of 10
logmodel.mml$`Predicted Total Cell Count` <- 10^as.numeric(logmodel.mml$`Predicted Total Cell Count`) 
logmodel.mml$`Lower Bound` <- 10^as.numeric(logmodel.mml$`Lower Bound`) #lower bound of cell counts
logmodel.mml$`Upper Bound` <- 10^as.numeric(logmodel.mml$`Upper Bound`) #upper bound of cell counts
logmodel.mml$RNA.ng.µL <- 10^(logmodel.mml$log10_RNA.ng.µL - 1) #substract 1 to undo the shift transformation and obtain the RNA conc.

#define a new variable for the predicted cell counts from this study
ThisStudy_PredCellCounts <- NA

#take the absolute value of the log10 transformation of [RNA]
log.RNA_thisStudy <- abs(log10(RNA.2.Biomass_TableS10$`RNA (ng/µL)`))

#now predict the lower bound for cell concentration from this study based on [RNA]
for (i in 1:length(log.RNA_thisStudy)){
  ThisStudy_PredCellCounts[i] <- (10^Cmax)*(log.RNA_thisStudy[i]/
                                         (10^Km + log.RNA_thisStudy[i]))
}

#Append results to RNA.2.Biomass_TableS10 data frame
RNA.2.Biomass_TableS10$PredCellCounts <- ThisStudy_PredCellCounts


```


```{r, merge data frames and obtain statistics}
#create a master data frame that encompasses both Table_S9 and Table_S10 data
RNA.2.Biomass <- rbind(RNA.2.Biomass_TableS9, RNA.2.Biomass_TableS10)

#add a column that presents known cell counts at experiment end from Table S9 and predicted cell counts from Table S10
RNA.2.Biomass$`Act. + pred. Num. cells at exp end` <- NA
#Table S9 direct cell count values (Harris et al., 2021) 
RNA.2.Biomass$`Act. + pred. Num. cells at exp end`[1:30] <- RNA.2.Biomass_TableS9$`Total Direct Cell Counts (AO) Experiment End`[1:30]
#Table S10 predicted cell count values (This Study)
RNA.2.Biomass$`Act. + pred. Num. cells at exp end`[31:60] <- RNA.2.Biomass_TableS10$PredCellCounts[1:30]

#create a molten data frame from selected values
melt.RNA.2.Biomass <- melt(RNA.2.Biomass, id.vars = c("Study", "Temperature (˚C)", "Condition", "Pressure (mbar)", "Headspace",
                                                      "Replicate", "RNA (ng/µL)"),
                       measure.vars = c("Act. + pred. Num. cells at exp end"))

# summarySE provides the standard deviation, standard error of the mean, and a (default 95%) confidence interval
summary_pred <- summarySE(melt.RNA.2.Biomass, measurevar = c("value"), groupvars = c("Study", "`Temperature (˚C)`", "Condition", "`Pressure (mbar)`", "Headspace"))
summary_pred$method <- "model.prediction"
```

```{r, plot model fit}
plot.logmodel.mml <- ggplot(logmodel.mml, aes(x = RNA.ng.µL, y = as.numeric(`Predicted Total Cell Count`)/10))+
  geom_line(color = "grey80")+
  geom_ribbon(aes(ymin = as.numeric(`Predicted Total Cell Count`)/30 - as.numeric(`Lower Bound`)/30, #minimum is divided by 30 to account for 30 mL incubations
                    ymax = as.numeric(`Predicted Total Cell Count`)/10 + as.numeric(`Upper Bound`)/10, alpha = 0.25))+ #maximum is divided by 10 to account for 10 mL incubations
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)), limits = c(10^5, 10^9), "Cells/mL at Experiment End")+
  xlab("RNA Recovery (ng/µL)")+
  scale_x_continuous(limits = c(0, 30))+
  scale_shape_manual(values=c(21, 22))+
  scale_color_manual(values = c("#CC3300","black"))+
  scale_fill_manual(values = c("#66C2A5", "#FC8D62"))+
  geom_point(data = RNA.2.Biomass[1:30,], aes(x = `RNA (ng/µL)`, y = `AO-based Total Cells @ T = 28`, #This study
                                       color = factor(`Pressure (atm)`),
                                       shape = factor(`Temperature (˚C)`)), fill = "white",
             size = 3, stroke = 1.1)+
  geom_point(data = RNA.2.Biomass[31:60,], aes(x = `RNA (ng/µL)`, y = `Act. + pred. Num. cells at exp end`/`Volume of media (mL)`, #Harris 2021 data points
                                       color = factor(`Pressure (atm)`),
                                       shape = factor(`Temperature (˚C)`), fill = factor(Headspace)), 
             size = 3, stroke = 1.1)+
  theme_base()+
  facet_zoom(xlim = c(0.5, 2.5), ylim = c(1e6, 1e8))+
  theme(legend.position = "none")
  
suppressWarnings(print(plot.logmodel.mml))
